Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-04-04 paper.qmd
Updated Figure 14

# umap_pbmc <- read_rds("data/pbmc3k/pbmc_umap_12_min_dist_0.99.rds")
umap_pbmc <- read_rds("data/pbmc3k/pbmc_umap_3_min_dist_0.6.rds")
nldr2 <- umap_pbmc |>
  ggplot(aes(x = emb1,
             y = emb2))+
  geom_point(alpha=0.1, size=1, colour=clr_choice) +
  interior_annotation("b")

@@

<!--Fit the best model for author suggestion and compute error-->
```{r}
#| label: hexbin-pbmc
## Compute hexbin parameters
num_bins_x_pbmc <- 60

@@

df_model_data_pbmc_n <- bind_rows(df_b_pbmc, data_pbmc)

langevitour::langevitour(df_model_data_pbmc_n[1:(length(df_model_data_pbmc_n)-1)],
                         lineFrom = distance_df_small_edges_pbmc$from,
                         lineTo = distance_df_small_edges_pbmc$to,
                         group = factor(df_model_data_pbmc_n$type,
                                        c("data", "model")),
                         levelColors = c("#a65628", "#000000"))
```

@@

pbmc_proj_umap_model1 <- plot_proj(
  proj_obj = proj_obj1, 
  point_param = c(1.5, 0.2, "#a65628"), # size, alpha, color
  plot_limits = c(-0.65, 0.65), 
  title = "a2", 
  cex = 2, 
  axis_text_size = 5,
  is_color = FALSE)

@@

pbmc_proj_umap_model2 <- plot_proj(
    proj_obj = proj_obj2, 
    point_param = c(1.5, 0.2, "#a65628"), # size, alpha, color
    plot_limits = c(-1, 0.75), 
    title = "a3", 
    cex = 2, 
    axis_text_size = 5,
    is_color = FALSE)

@@

error_pbmc_umap <- read_rds("data/pbmc3k/error_pbmc_umap_30_min_dist_0.3.rds")
error_pbmc_umap2 <- read_rds("data/pbmc3k/error_pbmc_umap_5_min_dist_0.01.rds")
# error_pbmc_umap3 <- read_rds("data/pbmc3k/error_pbmc_umap_12_min_dist_0.99.rds")
error_pbmc_umap3 <- read_rds("data/pbmc3k/error_pbmc_umap_3_min_dist_0.6.rds")
error_pbmc_tsne <- read_rds("data/pbmc3k/error_pbmc_tsne_5.rds")
error_pbmc_tsne2 <- read_rds("data/pbmc3k/error_pbmc_tsne_30.rds")
error_pbmc_phate <- read_rds("data/pbmc3k/error_pbmc_phate_5.rds")
error_pbmc_trimap <- read_rds("data/pbmc3k/error_pbmc_trimap_12_4_3.rds")
error_pbmc_pacmap <- read_rds("data/pbmc3k/error_pbmc_pacmap_30_random_0.9_5.rds")

@@

error_pbmc <- error_pbmc |>
  mutate(method = factor(method,
                         levels = c("UMAP_3_min_dist_0.6", "UMAP_30_min_dist_0.3", "UMAP_5_min_dist_0.01","tsne_5", "tsne_30", "phate_5", "trimap_12_4_3", "pacmap")))
```

@@

```{r}
#| label: umap-pbmc-read-best
umap_pbmc2 <- read_rds("data/pbmc3k/pbmc_umap_3_min_dist_0.6.rds")
pbmc_scaled_obj <- gen_scaled_data(
  data = umap_pbmc2)
umap_pbmc_scaled_best <- pbmc_scaled_obj$scaled_nldr

@@

## Compute hexbin parameters
num_bins_x_pbmc <- 29

@@

@@

df_model_data_pbmc_n <- bind_rows(df_b_pbmc, data_pbmc)

langevitour::langevitour(df_model_data_pbmc_n[1:(length(df_model_data_pbmc_n)-1)],
                         lineFrom = distance_df_small_edges_pbmc$from,
                         lineTo = distance_df_small_edges_pbmc$to,
                         group = factor(df_model_data_pbmc_n$type,
                                        c("data", "model")),
                         levelColors = c("#999999", "#000000"))
```

@@

pbmc_proj_umap_model1_best <- plot_proj(
    proj_obj = proj_obj1,
    point_param = c(1.5, 0.2, "#999999"), # size, alpha, color
    plot_limits = c(-0.65, 0.65), 
    title = "b2", 
    cex = 2, 
    axis_text_size = 5,
    is_color = FALSE)

@@

pbmc_proj_umap_model2_best <- plot_proj(
    proj_obj = proj_obj2,
    point_param = c(1.5, 0.2, "#999999"), # size, alpha, color
    plot_limits = c(-1, 0.75), 
    title = "b3", 
    cex = 2, 
    axis_text_size = 5,
    is_color = FALSE)
```

>>>>

# umap_pbmc <- read_rds("data/pbmc3k/pbmc_umap_12_min_dist_0.99.rds")
umap_pbmc <- read_rds("data/pbmc3k/pbmc_umap_5_min_dist_0.8.rds")
nldr2 <- umap_pbmc |>
  ggplot(aes(x = emb1,
             y = emb2))+
  geom_point(alpha=0.1, size=1, colour=clr_choice) +
  interior_annotation("b")

@@

<!--Fit the best model for author suggestion and compute error-->
```{r}
#| label: hexbin-pbmc
## Compute hexbin parameters
num_bins_x_pbmc <- 26

@@

data_pbmc_n <- data_pbmc |>
  select(-type) |>
  mutate(type = as.character(umap_pbmc_scaled_with_cluster$cluster))

df_model_data_pbmc_n <- bind_rows(df_b_pbmc, data_pbmc_n)

langevitour::langevitour(df_model_data_pbmc_n[1:(length(df_model_data_pbmc_n)-1)],
                         lineFrom = distance_df_small_edges_pbmc$from,
                         lineTo = distance_df_small_edges_pbmc$to,
                         group = factor(df_model_data_pbmc_n$type,
                                        c("cluster1", "cluster2", "cluster3", "model")),
                         levelColors = c("#8dd3c7", "#fdb462", "#bebada", "#000000"))
```

@@

proj_obj1[["cluster"]] <- as.character(umap_pbmc_scaled_with_cluster$cluster)

pbmc_proj_umap_model1 <- plot_proj(
  proj_obj = proj_obj1, 
  point_param = c(1.5, 0.2), # size, alpha, color
  plot_limits = c(-0.65, 0.65), 
  title = "a2", 
  cex = 2, 
  axis_text_size = 5,
  is_color = TRUE) + scale_color_manual(values = c("#8dd3c7", "#fdb462", "#bebada"))

@@

proj_obj2[["cluster"]] <- as.character(umap_pbmc_scaled_with_cluster$cluster)

pbmc_proj_umap_model2 <- plot_proj(
    proj_obj = proj_obj2, 
    point_param = c(1.5, 0.2), # size, alpha, color
    plot_limits = c(-1, 0.75), 
    title = "a3", 
    cex = 2, 
    axis_text_size = 5,
    is_color = TRUE) + scale_color_manual(values = c("#8dd3c7", "#fdb462", "#bebada"))

@@

error_pbmc_umap <- read_rds("data/pbmc3k/error_pbmc_umap_30_min_dist_0.3.rds")
error_pbmc_umap2 <- read_rds("data/pbmc3k/error_pbmc_umap_5_min_dist_0.01.rds")
# error_pbmc_umap3 <- read_rds("data/pbmc3k/error_pbmc_umap_12_min_dist_0.99.rds")
# error_pbmc_umap3 <- read_rds("data/pbmc3k/error_pbmc_umap_3_min_dist_0.6.rds")
# error_pbmc_umap3 <- read_rds("data/pbmc3k/error_pbmc_umap_4_min_dist_0.6.rds")
# error_pbmc_umap3 <- read_rds("data/pbmc3k/error_pbmc_umap_5_min_dist_0.8.rds")
error_pbmc_umap3 <- read_rds("data/pbmc3k/error_pbmc_umap_5_min_dist_0.8.rds")
error_pbmc_tsne <- read_rds("data/pbmc3k/error_pbmc_tsne_5.rds")
error_pbmc_tsne2 <- read_rds("data/pbmc3k/error_pbmc_tsne_30.rds")
error_pbmc_phate <- read_rds("data/pbmc3k/error_pbmc_phate_5.rds")
error_pbmc_trimap <- read_rds("data/pbmc3k/error_pbmc_trimap_12_4_3.rds")
error_pbmc_pacmap <- read_rds("data/pbmc3k/error_pbmc_pacmap_30_random_0.9_5.rds")

@@

error_pbmc <- error_pbmc |>
  mutate(method = factor(method,
                         levels = c("UMAP_5_min_dist_0.8", "UMAP_30_min_dist_0.3", "UMAP_5_min_dist_0.01","tsne_5", "tsne_30", "phate_5", "trimap_12_4_3", "pacmap")))
```

@@

```{r}
#| label: umap-pbmc-read-best
umap_pbmc2 <- read_rds("data/pbmc3k/pbmc_umap_5_min_dist_0.8.rds")
pbmc_scaled_obj <- gen_scaled_data(
  data = umap_pbmc2)
umap_pbmc_scaled_best <- pbmc_scaled_obj$scaled_nldr

@@

## Compute hexbin parameters
num_bins_x_pbmc <- 18

@@

distance_df_small_edges_pbmc <- distance_pbmc |>
  filter(distance < 0.07) #benchmark_pbmc

tr_from_to_df_pbmc <- right_join(
  tr_from_to_df_pbmc, distance_df_small_edges_pbmc,
  by = c("from", "to"))

@@

df_model_data_pbmc_n <- bind_rows(df_b_pbmc, data_pbmc_n)

langevitour::langevitour(df_model_data_pbmc_n[1:(length(df_model_data_pbmc_n)-1)],
                         lineFrom = distance_df_small_edges_pbmc$from,
                         lineTo = distance_df_small_edges_pbmc$to,
                         group = factor(df_model_data_pbmc_n$type,
                                        c("cluster1", "cluster2", "cluster3", "model")),
                         levelColors = c("#8dd3c7", "#fdb462", "#bebada", "#000000"))
```

@@

proj_obj1[["cluster"]] <- as.character(umap_pbmc_scaled_with_cluster$cluster)

pbmc_proj_umap_model1_best <- plot_proj(
    proj_obj = proj_obj1,
    point_param = c(1.5, 0.2), # size, alpha, color
    plot_limits = c(-0.65, 0.65), 
    title = "b2", 
    cex = 2, 
    axis_text_size = 5,
    is_color = TRUE) + scale_color_manual(values = c("#8dd3c7", "#fdb462", "#bebada"))

@@

proj_obj2[["cluster"]] <- as.character(umap_pbmc_scaled_with_cluster$cluster)

pbmc_proj_umap_model2_best <- plot_proj(
    proj_obj = proj_obj2,
    point_param = c(1.5, 0.2), # size, alpha, color
    plot_limits = c(-1, 0.75), 
    title = "b3", 
    cex = 2, 
    axis_text_size = 5,
    is_color = TRUE) + scale_color_manual(values = c("#8dd3c7", "#fdb462", "#bebada"))
```
